A theorical-practical approach to manifold learning

### 2018-11-19 DimReduction/Mds/Documentation/Article_mds.tex
Adding s curve dataset to mds
Finishing Mds article

\begin{abstract}
    This work presents an theorical and practical approximation to the problem
    of non-linear dimensionality reduction using the multidimensional scaling
    method and provide an algorithmic implementation easy and understandable. In
    this approach a distance matrix $D^x$ is constructed to approximate the
    distance over a manifold by solving shortest path problems over a graph that
    represents a multidimensional structure. This matrix is used as input to the
    method, which is tested with 3 dimensional manifolds. This technique gives a
    way to improve the results using other ways to calculate the distance matrix.
\end{abstract}

@@

Some of the approximations in these types of methods assume that the data that
is being analyzed does not exploit all the dimensionality of the space where
is defined. Due to this fact, another description of the goal that the
dimensionality reduction techniques have is related to find a meaningful low
dimensional structure that is hidden in a multidimensional structure.
\cite{dimension}

@@

A clear example of these type of methods are manifold learning techniques. These
approximations search for a way to perform a non linear dimensionality reduction,
which means that they do not assume that the structure of the original dataset
is encoded based on euclidean distance meassuares. Instead these algorithms try
to express their relationships in terms of geodesic distances, which are
distances meassured over the manifold surface. The result of this intend is
to preserve the intrinsic structure of the data when the reduction is perfomed,
allowing to conserve the implicit information that was hidden in the original
structure of the dataset. \cite{manifold}.

@@

This work aims to develop the theorical and practical aspects of the method
described above, with the objective of build a method that can reduce a
multidimensional manifold to a low dimensional representation. To achieve this
in the first part of the work a theorical framework is developed, in order to
understand all the aspects of the method. Then an implementation section will
take place, followed by a result section. Later, some conclusions are
pressented and finally an apendix is added to present some proofs of the
steps that are used in the theorical framework.

@@

    As mentioned before, one of the goals of this work is to create a theorical
    development of the multidimensional scaling algorithm. This section shows a
    detail explanation of the therical aspects of the method taking \cite{proof}
    as base point.

@@

    Now, using that knowledge, the main expression can be expanded

    \begin{equation*}
        \begin{aligned}
        Tr[(\Lambda - G \hat{\Lambda} G^t)^2]
        &=
        Tr[\Lambda^2 - 2 \Lambda G \hat{\Lambda} G^t + (G \hat{\Lambda} G^t)^2]\\
        =
        &Tr[\Lambda^2 - 2 \Lambda G \hat{\Lambda} G^t + G \hat{\Lambda} G^t G \hat{\Lambda} G^t]\\
        \end{aligned}
    \end{equation*}

    But, knowing that $Q^tQ = I = QQ^t$, the following properties can be developed

    \begin{equation*}
        \begin{aligned}
            G^tG &= (V^tQ)^t V^tQ\\
                 &= (Q^tV) V^tQ\\
                 &= Q^tQ\\
                 &= I
        \end{aligned}
    \end{equation*}

    \begin{equation*}
        \begin{aligned}
            GG^t &= V^tQ (V^tQ)^t\\
                 &= V^tQ (Q^tV)\\
                 &= V^tV\\
                 &= I\\ \\
        \end{aligned}
    \end{equation*}

    So, the following transformations can be made to the problem

    \begin{equation*}
        \begin{aligned}
        Tr[(\Lambda - G \hat{\Lambda} G^t G)^2 ]
        &=
        Tr[\Lambda^2 - 2 \Lambda G \hat{\Lambda} G^t + G \hat{\Lambda} \hat{\Lambda} G^t]\\
        =
        Tr[&\Lambda^2 - 2 \Lambda G \hat{\Lambda} G^t + G \hat{\Lambda}^2 G^t]\\
        =
        Tr[&\Lambda^2]- 2 Tr[\Lambda G \hat{\Lambda} G^t] + Tr[G \hat{\Lambda}^2 G^t]\\
        =
        Tr[&\Lambda^2] - 2 Tr[\Lambda G \hat{\Lambda} G^t] + Tr[ \hat{\Lambda}^2 G^tG]\\
        =
        Tr[&\Lambda^2] - 2 Tr[\Lambda G \hat{\Lambda} G^t] + Tr[ \hat{\Lambda}^2]\\
        =
        Tr[&\Lambda^2] + Tr[ \hat{\Lambda}^2] - 2 Tr[\Lambda G \hat{\Lambda} G^t]\\
        \end{aligned}
    \end{equation*}

    Which takes us to the \textbf{DARK PART!}\\

    We know that $\frac{\delta Tr[\Lambda G \hat{\Lambda} G^t]}{\delta G} = 2\Lambda G \hat{\Lambda}$\\\\
    We also know that $\frac{\delta Tr(\Lambda^2)}{\delta G} = 0 = \frac{\delta Tr(\hat{\Lambda}^2)}{\delta G}$\\\\
    Finally we know that the answer for a fixed $\hat{\Lambda}$ is $G=I$\\

    \textbf{END OF DARK PART}\\

@@

    To minimize the resulting expression, the first $p$ eigenvalues of $\Lambda$
    can be used to define $\hat{\Lambda}$. And finally the solution to the
    problem can be found by the following procedure

@@

    which implies that $Y = \hat{\Lambda}^{1/2} V^t$. Which solves the problem.

@@

    To solve the problem the first step is rewrite each entry of matrix $D$ in
    a vector form

@@

The method described in the previuos section was tested with 3 dimensional
manifolds in order to allow the visualization of the intial manifold and be able
to compare the quality of the resulting solution. Figure \ref{swiss} shows
a swiss roll dataset that was used as the multidimensional manifold. Figure
\ref{swiss_res} shows the result of the devolped method.

@@

\begin{figure}[H]
    \centering
    \includegraphics[width=0.7\linewidth]{images/swiss_roll_result.png}
    \caption{Swiss roll dataset reduced}
    \label{swiss_res}
\end{figure}

@@

In this work a theorical and practical development of the multidimensional
scaling method was permed. During the process it was found that what seemed like
a fairly complicated problem, can be simplified to simple matrix operations
with the help of some concepts from linear algebra. Thanks to that theorical
framework, the construction of a relatively simple algorithm that can detect
low dimensional manifolds embedded in multidimensional spaces was possible.
Finally the results shown that this technique has the ability to make
contributions to the analysis proces of multidimensional datasets by
simplifying its features but preserving the intrinsic structure.

@@

    And if $Q$ is also sqare, then

@@

    So, if a new matrix $\Lambda$ is define as a diagonal matrix that contains
    the eigenvalues of $A$, then the computation can be expressed as

>>>>

\begin{abstract}
    This work presents an theorical and practical approximation to the problem
    of non-linear dimensionality reduction using the multidimensional scaling
    method and provides an algorithmic implementation easy and understandable. In
    this approach a distance matrix $D^x$ is constructed to approximate the
    distance over a manifold by solving shortest path problems over a graph that
    represents a multidimensional structure. This matrix is used as input to the
    method, which is tested with 3 dimensional manifolds. This technique gives a
    way to improve the results using other ways to calculate the distance matrix.
\end{abstract}

@@

Some of the approximations in these types of methods assume that the data that
is being analyzed does not exploit all the dimensionality of the space where
it is defined. Due to this fact, another description of the goal that the
dimensionality reduction techniques have is related to find a meaningful low
dimensional structure that is hidden in a multidimensional structure.
\cite{dimension}

@@

A clear example of these type of methods are manifold learning techniques. These
approximations search for a way to perform a non linear dimensionality reduction,
which means that they do not assume that the structure of the original dataset
is encoded based on euclidean distance meassures. Instead these algorithms try
to express their relationships in terms of geodesic distances, which are
distances meassured over the manifold surface. The result of this intend is
to preserve the intrinsic structure of the data when the reduction is perfomed,
allowing to conserve the implicit information that was hidden in the original
structure of the dataset. \cite{manifold}.

@@

This work aims to develop the theorical and practical aspects of the method
described above, with the objective of build a method that can reduce a
multidimensional manifold to a low dimensional representation. To achieve this
in the first part of the work a theorical framework is developed in order to
understand all the aspects of the method. Then an implementation section will
take place, followed by a result section. Later, some conclusions are
pressented and finally an apendix is added to present some proofs of the
steps that are used in the theorical framework.

@@

    As mentioned before, one of the goals of this work is to create a theorical
    development of the multidimensional scaling algorithm. This section shows a
    detail explanation of the theorical aspects of the method taking \cite{proof}
    as base point.

@@

    Now, since this is a quadratic expression, a minimum can be found by making
    $\Lambda$ as close as possible to $G \hat{\Lambda} G^t$. Then $G$ must be
    an identity matrix.\\

%    Now, using that knowledge, the main expression can be expanded
%
%    \begin{equation*}
%        \begin{aligned}
%        Tr[(\Lambda - G \hat{\Lambda} G^t)^2]
%        &=
%        Tr[\Lambda^2 - 2 \Lambda G \hat{\Lambda} G^t + (G \hat{\Lambda} G^t)^2]\\
%        =
%        &Tr[\Lambda^2 - 2 \Lambda G \hat{\Lambda} G^t + G \hat{\Lambda} G^t G \hat{\Lambda} G^t]\\
%        \end{aligned}
%    \end{equation*}
%
%    But, knowing that $Q^tQ = I = QQ^t$, the following properties can be developed
%
%    \begin{equation*}
%        \begin{aligned}
%            G^tG &= (V^tQ)^t V^tQ\\
%                 &= (Q^tV) V^tQ\\
%                 &= Q^tQ\\
%                 &= I
%        \end{aligned}
%    \end{equation*}
%
%    \begin{equation*}
%        \begin{aligned}
%            GG^t &= V^tQ (V^tQ)^t\\
%                 &= V^tQ (Q^tV)\\
%                 &= V^tV\\
%                 &= I\\ \\
%        \end{aligned}
%    \end{equation*}
%
%    So, the following transformations can be made to the problem
%
%    \begin{equation*}
%        \begin{aligned}
%        Tr[(\Lambda - G \hat{\Lambda} G^t G)^2 ]
%        &=
%        Tr[\Lambda^2 - 2 \Lambda G \hat{\Lambda} G^t + G \hat{\Lambda} \hat{\Lambda} G^t]\\
%        =
%        Tr[&\Lambda^2 - 2 \Lambda G \hat{\Lambda} G^t + G \hat{\Lambda}^2 G^t]\\
%        =
%        Tr[&\Lambda^2]- 2 Tr[\Lambda G \hat{\Lambda} G^t] + Tr[G \hat{\Lambda}^2 G^t]\\
%        =
%        Tr[&\Lambda^2] - 2 Tr[\Lambda G \hat{\Lambda} G^t] + Tr[ \hat{\Lambda}^2 G^tG]\\
%        =
%        Tr[&\Lambda^2] - 2 Tr[\Lambda G \hat{\Lambda} G^t] + Tr[ \hat{\Lambda}^2]\\
%        =
%        Tr[&\Lambda^2] + Tr[ \hat{\Lambda}^2] - 2 Tr[\Lambda G \hat{\Lambda} G^t]\\
%        \end{aligned}
%    \end{equation*}
%
%    Which takes us to the \textbf{DARK PART!}\\
%
%    We know that $\frac{\delta Tr[\Lambda G \hat{\Lambda} G^t]}{\delta G} = 2\Lambda G \hat{\Lambda}$\\\\
%    We also know that $\frac{\delta Tr(\Lambda^2)}{\delta G} = 0 = \frac{\delta Tr(\hat{\Lambda}^2)}{\delta G}$\\\\
%    Finally we know that the answer for a fixed $\hat{\Lambda}$ is $G=I$\\
%
%    \textbf{END OF DARK PART}\\

@@

    To minimize the resulting expression, the first $p$ eigenvalues of $\Lambda$
    can be used to define $\hat{\Lambda}$. Finally, the solution to the
    problem can be found by the following procedure

@@

    which implies that $Y = \hat{\Lambda}^{1/2} V^t$. And that solves the problem.

@@

    To solve this problem the first step is rewrite each entry of matrix $D$ in
    a vector form

@@

The method described in the previuos section was tested with 3 dimensional
manifolds in order to allow the visualization of the intial manifold and be able
to compare the quality of the resulting solution. Figure \ref{swiss} shows
a swiss roll dataset that was used as the multidimensional manifold. Figure
\ref{swiss_res} shows the result of the devolped method. Finally in Figure       
\ref{s_curve} there is a s curve dataset and Figure \ref{s_curve_res} shows its  
corresponding simplified output.

@@

\begin{figure}[H]
    \centering
    \includegraphics[width=0.6\linewidth]{images/swiss_roll_result.png}
    \caption{Swiss roll dataset reduced}
    \label{swiss_res}
\end{figure}

\begin{figure}[H]                                                                
    \centering                                                                   
    \includegraphics[width=0.9\linewidth]{images/s_curve.png}                    
    \caption{S curve dataset}                                                    
    \label{s_curve}                                                              
\end{figure}

\begin{figure}[H]                                                                
    \centering                                                                   
    \includegraphics[width=0.9\linewidth]{images/s_curve_result.png}             
    \caption{S curve dataset reduced}                                            
    \label{s_curve_res}                                                          
\end{figure}

@@

In this work a theorical and practical development of the multidimensional
scaling method was conducted. During the process it was found that what seemed like
a fairly complicated problem, can be simplified to simple matrix operations
with the help of some concepts from linear algebra. Thanks to that theorical
framework, the construction of a relatively simple algorithm that can detect
low dimensional manifolds embedded in multidimensional spaces became possible.
The results shown that this technique has the ability to make
contributions to the analysis proces of multidimensional datasets by
simplifying its features but preserving the intrinsic structure. Finally, as
future work, this method can be improved by creating new ways to compute the
matrix $D^x$. It would be interesting to implement this method using GPU
computations to improve its performance and use it as a mechanism of
data preprocessing in machine learning algorithms.

@@

    And if $Q$ is also square, then

@@

    So, if a new matrix $\Lambda$ is defined as a diagonal matrix that contains
    the eigenvalues of $A$, then the computation can be expressed as
